Automatically Splitting a Two-Stage Lambda Calculus

### 2016-01-06 paper/intro.tex
Carlo's comments

While simple forms of frequency reduction include standard compiler
optimizations such as loop hoisting and common subexpression
elimination, J{\o}rring and Scherlis proposed the more general
transformation of {\em splitting} a program into multiple subfunctions
(called pass separation in \,\cite{JS86-staging}). In our example,
splitting transforms the function $f$ into two others $f_1$ and $f_2$
such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating $f_1$ on $x$, and using the result $z$ to evaluate
the second function on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between splitting and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Therefore, unlike partial
evaluation, splitting is a static program transformation
(``metastatic'' in partial evaluation terms) and does not require
dynamic code generation.

>>>>

While simple forms of frequency reduction include standard compiler
optimizations such as loop hoisting and common subexpression
elimination, J{\o}rring and Scherlis proposed the more general
transformation of {\em splitting} a program into multiple subfunctions
(called pass separation in \,\cite{JS86-staging}). In our example,
splitting transforms the function $f$ into two others $f_1$ and $f_2$
such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating $f_1$ on $x$, and using the result $z$ to evaluate
the second function on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between splitting and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Therefore, unlike partial
evaluation, splitting is a static program transformation
(``metastatic'' in partial evaluation terminology) and does not require
dynamic code generation.

### 2016-01-06 paper/semantics.tex
Carlo's comments

We express two-stage programs as terms in \lang, a typed, modal lambda calculus. 
Although \lang\ describes computations that occur in two stages,
we find it helpful for the specification of splitting to differentiate terms into three \emph{worlds}.
A world is essentially a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within {\tt next} blocks,
and \bbonep\ for \emph{ground} first-stage computations, which may not.
The distinction between these two first-stage worlds is necessary for an efficient splitting algorithm
and will be discussed in \ref{sec:needGround}.

@@

All three worlds contain unit, product, function, sum and recursive types defined in the usual fashion.
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.
Thus differing worlds (and hence, differing stages of computation) only interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of typing at worlds \bbtwo\ and \bbonep, respectively, to typing at \bbonem.

@@

Consider the following example, where {\tt P} has type $\curr(\rmunit + \rmunit)$,\begin{lstlisting}
           1`caseg P of _ => next{`2`0`1`} | _ => next{`2`literalone`1`}`
\end{lstlisting}
Depending on what {\tt P} evaluates to in the first stage, whole term will step to either {\tt next\{0\}} or {\tt next\{1\}}.
In this sense, we can see {\tt case} (at world \bbonem) and {\tt caseg} as the constructs that facilitate
all cross-stage communication.

>>>>

We express two-stage programs as terms in \lang, a typed, modal lambda calculus. 
Although \lang\ describes computations that occur in two stages,
we find it helpful for the specification of splitting to codify terms using one of three \emph{worlds}.
A world is essentially a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within {\tt next} blocks,
and \bbonep\ for \emph{ground} first-stage computations, which may not.
The distinction between these two first-stage worlds is necessary 
for the splitting algorithm to produce efficient outputs
and will be discussed in \ref{sec:needGround}.

@@

All three worlds contain unit, product, function, sum and recursive types defined in the usual fashion.
These ``standard'' features can only be constructed from subterms of the same world,
and variables can only be used at the same world where they were introduced.
Thus differing worlds (and hence, differing stages of computation) only interact by means
of the $\fut$ and $\curr$ type formers.  These modalities are internalizations
of worlds \bbtwo\ and \bbonep, respectively, to types at world \bbonem.

@@

Consider the following example, where {\tt P} has type $\curr(\rmunit + \rmunit)$,\begin{lstlisting}
           1`caseg P of _ => next{`2`0`1`} | _ => next{`2`literalone`1`}`
\end{lstlisting}
Depending on what {\tt P} evaluates to in the first stage, 
the whole term will step to either {\tt next\{0\}} or {\tt next\{1\}}.
In this sense, we can see {\tt case} (at world \bbonem) and {\tt caseg} as the constructs that facilitate
all cross-stage communication.

### 2016-01-06 paper/splitting.tex
Carlo's comments

Splitting a partial value $v$, via $\vsplito$, yields $\mval i q$.
Since partial values, by definition, have no remaining work in the first pass
and since transfer of information between the stages occurs in the first pass of evalution,
we know that the second stage components of $v$ can no longer depend on its first stage components.
Analogously, this must also hold for the output of splitting, 
which is why $q$---unlike the resumer of world \bbtwo\ term splitting---is not open on a variable.
Thus, $i$ and $q$ are operationally independent, but they represent the complementary portions of $v$ that are relevant to each stage.
We call $i$ the \emph{immediate value} and $q$ the residual.

@@

This structure can be described with a type; for instance, the staged quickselect yields a binary search tree.
Indeed, for most of our examples in \ref{sec:examples}, inferring this type is straight-forward. 
However, we do not yet have a formal characterization of these boundary types that is defined for all \lang\ programs,
though we plan to pursue this in future work.
\footnote{Observe that in the outputs of splitting we could have ommitted the {\tt roll} tag from functions or 
replaced $\inl a$ and $\inr b$ with $\tup {0,a}$ and $\tup {1,b}$, and the proof of correctness would still have gone through.
We chose the tags, however, in order to keep the typed interpretation more natural, even in absence of a formal result.
}
\end{abstrsyn}

>>>>

Splitting a partial value $v$, via $\vsplito$, yields $\mval i q$.
Since partial values, by definition, have no remaining work in the first pass
and since transfer of information between the stages occurs in the first pass of evalution,
we know that the second-stage components of $v$ can no longer depend on its first-stage components.
Analogously, this must also hold for the output of splitting, 
which is why $q$---unlike the resumer of world \bbtwo\ term splitting---is not open on a variable.
Thus, $i$ and $q$ are operationally independent, but they represent the complementary portions of $v$ that are relevant to each stage.
We call $i$ the \emph{immediate value} and $q$ the residual.

@@

This structure can be described with a type; for instance, the staged quickselect yields a binary search tree.
Indeed, for most of our examples in \ref{sec:examples}, inferring this type is straight-forward. 
However, we do not yet have a formal characterization of these boundary types that is defined for all \lang\ programs,
though we plan to pursue this in future work.
\footnote{Observe that in the outputs of splitting we could have omitted the {\tt roll} tag from functions or 
replaced $\inl a$ and $\inr b$ with $\tup {0,a}$ and $\tup {1,b}$, and the proof of correctness would still have gone through.
We chose the tags, however, in order to keep the typed interpretation more natural, even in absence of a formal result.
}
\end{abstrsyn}
